
---
title: "w04"
author: "Yekta Amirkhalili"
date: "May 29th, 2023"
output:
  html_document:
    df_print: paged
---


### Week 4 (Session 3) - May 29, 2023 
**material to cover:**

---
**0.**
...**Probability Distributions**...

All code references: https://cran.r-project.org/web/packages/Rlab/index.html

```{r}
library("Rlab") 

```

------------------
------------------

There are different types of random variables in Statistics. Here are some of the most well-known *Discrete* Random Variables (RV). 

1. Bernouli 
Bernoulli RV takes in values of "Success" and "Failure" which can be represented mathematically as 1 and 0 respectively. Let's define probability of success as $p$, and following that probability of failure would be $1-p$. The probability function would then be: 

$$ P(X = x) = \left\{\begin{aligned}
            &p && x = 1\\
            &1-p && x = 0
            \end{aligned}
            \right.$$
            
```{r, bernouli}
set.seed(111)                              
N <- 10000

random_vars_bern <- rbern(N, prob = 0.8)  
distro_bern <- dbern(random_vars_bern, 0.8, log = FALSE)
cdf_bern <- pbern(random_vars_bern, 0.8)

first_20 <- random_vars_bern[1:20]
print('The first 20 Bernoulli trials: ')
print(first_20)

print('The distribution')
print(distro_bern[1:20])

print('The CDF')
hist(cdf_bern)

```
        
            
2. Binomial 
The number of "Success"es in $n$ trials of independent Bernoulli experiments with success probability of $p$ follows a Binomial Distribution. The probability function is: 

$$\begin{equation} 
  P\left(X = i\right) = \binom{n}{i} p^i\left(1-p\right)^{n-i}
\end{equation}$$

Where $i = 0, 1, 2, ..., n$. 

```{r, binom}
set.seed(112)                              
N <- 10000

random_vars_binom <- rbinom(N, size = 10, prob = 0.7)  
distro_binom <- dbinom(random_vars_binom, size = 10, prob = 0.7)
cdf_binom <- pbinom(random_vars_binom, size = 10, prob = 0.7)

first_20 <- random_vars_binom[1:20]
print('The first 20 Binomial: ')
print(first_20)

print('The distribution')
print(distro_binom[1:20])

print('The CDF')
hist(cdf_binom)


```
3. Geometric 
The number of independent Bernoulli trials (with parameter $p$) to reach *the first* success follows a Geometric distribution. The probability function is: 

$$ P(X = i) = (1-p)^{i-1}p \:\:\:\:\:\:\:\:\ i = 1,2,3,... $$
```{r, geom}
set.seed(113)                              
N <- 10000

random_vars_geom <- rgeom(N, prob = 0.65)  
distro_geom <- dgeom(random_vars_geom, prob = 0.65)
cdf_geom <- pgeom(random_vars_geom, prob = 0.65)

first_20 <- random_vars_geom[1:20]
print('The first 20 Geometric: ')
print(first_20)

print('The distribution')
print(distro_geom[1:20])

print('The CDF')
hist(cdf_geom)

```


4. Negative Binomial 
The number of independent Benoulli trials (with parameter $p$) to reach the $r$th success follows a Negative Binomial distribution. The probability function is: 

$$\begin{equation} 
  P\left(X = i\right) = \binom{i-1}{r-1} p^r\left(1-p\right)^{i-r}
  \:\:\:\:\:\:\ i = r, r+1, r+2, ...
\end{equation}$$

```{r}
set.seed(114)                              
N <- 10000

random_vars_nb <- rnbinom(N, size = 9, prob = 0.7)
distro_nb <- dnbinom(random_vars_nb, size = 9, prob = 0.7)
cdf_nb <- pnbinom(random_vars_nb, size = 9, prob = 0.7)


first_20 <- random_vars_nb[1:20]
print('The first 20 Negative Binomial: ')
print(first_20)

print('The distribution')
print(distro_nb[1:20])

print('The CDF')
hist(cdf_nb)

```

5. Poisson
It is basically an approximation on Binomial Distribution when $n$ is a large number ($n \ge 20$) and the probability of success is small. The probability function is: 

$$ P(X = i) = \displaystyle \frac{e^{-\lambda}\lambda^i}{i!}$$

```{r}
set.seed(115)                              
N <- 100000

rv_pos <- rpois(N, 10)
dis_pos <- dpois(rv_pos, 10)
cdf_pos <- ppois(rv_pos, 10)

first_20 <- rv_pos[1:20]
print('The first 20 Poisson: ')
print(first_20)

print('The distribution')
print(dis_pos[1:20])

print('The CDF')
hist(cdf_pos)

```
6. Hyper Geometric 
Suppose there exists a box with $m$ defective and $N-m$ working parts. We randomly take out $n$ parts without replacement. If $X$ represents the number of defective parts taken out, we have: 

$$ P(X = i) = \displaystyle \frac{ \binom{m}{i}\binom{N-m}{n-i}}{\binom{N}{n}}$$

```{r}
set.seed(116)                              
N <- 10000
m <- 480
n <- N - m

rv_hyper <- rhyper(N, m, n, 5)
dis_hyper <- dhyper(rv_hyper, m, n, 5)
cdf_hyper <- phyper(rv_hyper, m, n, 5)

first_20 <- rv_hyper[1:20]
print('The first 20 Hyper Geometric: ')
print(first_20)

print('The distribution')
print(dis_hyper[1:20])

print('The CDF')
hist(cdf_hyper)


```


7. Uniform 
If a RV $X$ takes the values of $x_1, x_2, x_3, ..., x_n$ with equal chance of $\frac{1}{n}$, then $X$ has a uniform distribution with the following probability function: 

$$P(X = x_i) = \frac{1}{n}$$



```{r}
N <- 100

rv_unif <- rep(0.01, 100) #repeat 1/100 100 times
dis_unif <- function(nums){
    u <- 1/nums
    u_ <- rep(u, nums)
    
    return(u_)
} #number of values : 1/nums = probability 

dis_uniform <- dis_unif(N)

cdf_unif <- function(pr, l){
    cdf <- c()
    le <- length(l)
    
    for(i in le:1){
        cdf <- c(cdf, sum(l[1:i]))
    }
    
    return(rev(cdf))
}

cdf_uniform <- cdf_unif((1/N), rv_unif)

first_20 <- rv_unif[1:20]
print('The first 20 Uniform: ')
print(first_20)

print('The distribution')
print(dis_uniform[1:20])

print('The CDF')
hist(cdf_uniform)

```

More: Read up on packages that do this for you! 

Here are some of the most well-known *Continous* Random Variables (RV). 

1. Continuous Uniform 
The RV $X$ has a Uniform distribution in interval $(a,b)$ if the probability density function is: 

$$ f(x) = k = \frac{1}{b-a}, \:\:\:\:\:\ a \le x \le b $$
```{r, unif_cont}
set.seed(100)
N <- 1000

uni <- runif(N, min = 0, max = 100)
uni_dist <- dunif(uni, min = 0, max = 100, log = FALSE)
uni_cdf <- punif(uni, min = 0, max = 100) 


first_20 <- uni[1:20]
print('The first 20 Uniform: ')
print(first_20)

print('The distribution')
print(uni_dist[1:20])

plot(uni, uni_dist)

```




2. Normal 
It was first introduced as an approximation of Binomial (with parameters $n$ and $p=0.5$ where $n$ was large and $p$ was not too large and not too small). The probability density function is as follows: 


$$ f(x) = \frac{1}{\sigma \sqrt{2\pi}}e^{\frac{-(x-\mu)^2}{2\sigma^2}} \:\:\:\:\:\ -\infty \lt x \lt +\infty$$

```{r, norm}
set.seed(101)
N <- 1000

x <- rnorm(N, 0, 1)
y <- dnorm(x, mean = 0, sd = 1)

norm_cdf <- pnorm(x, 0, 1)

first_20 <- x[1:20]
print('The first 20 Normal: ')
print(first_20)

print('The distribution')
print(y[1:20])

plot(x, y)
```

3. Normal Approximation to the Binomial
If $X$ is a Binomial RV with parameters $n$ and $p$, and $n$ is large while $p$ is neither large nor small, then we can approximate this distribution with new parameters: $\mu = np$ and $\sigma^2 = np(1-p)$ and that would be a normal distribution. 


```{r, normapprx}

```

4. Exponential
RV $X$ follows the exponential distribution with parameter $\lambda \gt 0$ when it has the following probability density function: 

$$ f(x) = \lambda e^{-\lambda x} \:\:\:\ x \ge 0 $$
```{r, expo}
set.seed(105)
N <- 1000

e_ <- rexp(N, rate = 0.1)
e_dis <- dexp(e_, rate = 0.1)
e_cdf <- pexp(e_, rate = 0.1)

plot(e_, e_dis)
```

5. Gamma 
RV $X$ has a gamma distribution with parameters $\alpha \gt 0$ and $\lambda \gt 0$ if the probability density function is as follows: 

$$ f(x) = \frac{\lambda^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-\lambda x} \:\:\:\:\:\ x \ge 0$$
Read more about Gamma function: https://www.statlect.com/mathematical-tools/gamma-function

```{r, gamma}
set.seed(106)
N <- 1000

#shape is alpha, rate is lambda and scale is 1/lambda 
g <- rgamma(N, shape = 5, rate = 0.1)
dis_g <- dgamma(g, shape = 5, rate = 0.1)

plot(g, dis_g)
```

6. Beta
RV $X$ has a beta distribution with positive parameters $a$ and $b$ if the probability density function is: 

$$ f(x) = \frac{\Gamma(a+b)}{\Gamma(a) \Gamma(b)}x^{a-1}(1-x)^{b-1} \:\:\:\:\:\ 0 \lt x \lt 1 $$

```{r, beta}
set.seed(107)
N <- 1000

bet <- rbeta(N, shape1 = 2, shape2 = 7)
bet_dis <- dbeta(bet, shape1 = 2, shape2 = 7)

plot(bet, bet_dis)

```

Thought: What happens if $a = b = 1$ ? 

7. Chi-sqaured 

**NOTE**If a random variable $Z$ follows the Standard Normal distribution, then $Y = Z^2$ follows a gamma distribution with parameters $\lambda = 0.5$ and $\alpha = 0.5$. 

If $Z_1, Z_2, ..., Z_n$ are independent random variables following Standard Normal Distribution, then the RV $Y = Z_1^2 + Z_2^2 + ... + Z_n^2$ follows a gamma distribution with parameters $\lambda = 0.5$ and $\alpha = \frac{n}{2}$. This special case of Gamma is called a Chi-squared distribution with $n$ degrees of freedom, denoted by $\chi_n^2$.

```{r, chi}
set.seed(108)
N <- 1000

ci <- rchisq(N, 10)
ci_dis <- dchisq(ci, 10)

plot(ci, ci_dis)


```





Additional: LogNormal 
Read: https://www.itl.nist.gov/div898/handbook/eda/section3/eda3669.htm



